---
title: "main"
output: html_document
date: "2023-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
suppressMessages(library(tidymodels))
library(reshape2)
```

# This is a very straight forward dataset. I think we are just going to do a little EDA and a simple linear model.

```{r EDA}
artists <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-01-17/artists.csv', show_col_types = FALSE)

show(artists)

artists %>% count(artist_nationality, artist_gender)

artists %>% count(artist_race)

artists %>% count(year)

# Convert to actual NA's
artists[artists == "N/A"] <- NA 
artists <- artists %>% drop_na()
```

```{r EDA}
artists %>%
  drop_na() %>%
  ggplot(aes(year, fill = artist_race)) +
  geom_bar()

```

```{r EDA}
artists %>%
  drop_na() %>%
  ggplot(aes(year, fill = artist_gender)) +
  geom_bar()
```

```{r EDA}
artists %>%
  drop_na() %>%
  ggplot(aes(x = space_ratio_per_page_total, fill = artist_gender)) + 
  geom_histogram(bins = 15) +
  facet_wrap(vars(artist_gender), scales = "free_y", ncol = 1) +
  scale_x_log10() +
  labs(fill = NULL)
```

```{r Modeling}
linearModel <- lm(space_ratio_per_page_total ~ artist_gender + year,
                   data = artists)
summary(linearModel)
```
# despite our model not being particularly good, we would fail to reject our null hypothesis (i.e. men being alloted more area in books and that said area is expanding).
```{r Modeling}
set.seed(0)
boots <- bootstraps(artists, times = 1000, apparent = TRUE)
boots
```

```{r Modeling}
bootModels <-
  boots %>%
  mutate(
    model = map(
      splits,
      ~ lm(space_ratio_per_page_total ~ artist_gender + year,
           data = analysis(.x))
    ),
    coef_info = map(model, tidy))
```

```{r Modeling}
bootCoefs <- 
  bootModels %>% 
  unnest(coef_info)
```

```{r}
percentile_intervals <- int_pctl(bootModels, coef_info)

ggplot(bootCoefs, aes(estimate)) +
  geom_histogram(bins = 30) +
  facet_wrap( ~ term, scales = "free") +
  geom_vline(aes(xintercept = .lower), data = percentile_intervals, col = "blue") +
  geom_vline(aes(xintercept = .upper), data = percentile_intervals, col = "blue")
```
```{r Modeling}
bootModels %>%
  sample_n(150) %>%
  mutate(augmented = map(model,augment)) %>%
  unnest(augmented) %>%
  ggplot(aes(year, .fitted, color = artist_gender, group = id)) + geom_line(alpha =.2, show.legend = FALSE) +
  geom_line(alpha = 0.1, show.legend = FALSE) +
  facet_wrap(vars(artist_gender))
```
# In hindsight I should have transformed the data or tried a different model to get something more accurate.







